---
title: "An In-Depth Analysis of Temporal, Demographic, and Spatial Trends in NYPD Shooting Incident Data"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
author: "Mohamed Ghassen Brahim"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

# Introduction

This report aims to analyze the NYPD Shooting Incident Data to uncover trends and insights into shooting incidents over the years.
We will explore various aspects such as temporal trends, geographical distribution, and other relevant factors.

# Setup and Reproducibility

In order to be sure that all required dependency are installed, i recommend this lignes of code, which checks if the packages are installed otherwise it install them for us.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load necessary libraries (install them if not already installed)
libraries <- c("readr", "dplyr", "ggplot2", "tidyverse", "lubridate", "forecast", "tseries", "zoo", "MASS")
lapply(libraries, function(x) {
    if (!require(x, character.only = TRUE)) {
        install.packages(x)
        library(x, character.only = TRUE)
    }
})
```

# Data Importing

```{r}
# Load the NYPD Shooting Incident Data
raw_data <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```

# Data Tidying and Exploration

How Does our Data look like?

```{r}
# Summarize
summary(raw_data)
```

As we can see, we have 27312 entries.These includes 21 columns.

The Key Columns are:

-   INCIDENT_KEY
-   OCCUR_DATE
-   OCCUR_TIME
-   BORO (borough)
-   PRECINCT
-   Various perpetrator and victim details:
    -   age
    -   sex
    -   race
-   Latitude and Longitude : coordinates of the incidents
-   STATISTICAL_MURDER_FLAG: a flag indicating whether the incident was a statistical murder

# Data Transformation

The following columns are converted to the appropriate type (also provided).

| Column     | Original Data Type | Converted Data Type |
|------------|--------------------|---------------------|
| OCCUR_DATE | chr                | date                |
| OCCUR_TIME | chr                | factor              |
| BORO       | chr                | factor              |
| PRECINCT   | chr                | factor              |

```{r}
raw_data <- raw_data %>% 
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),
         OCCUR_TIME = as.factor(OCCUR_TIME),
         BORO = as.factor(BORO),
         PRECINCT = as.factor(PRECINCT))

```

Let's take a look at the data after transformations:

```{r}
# Summarize
summary(raw_data)
```

# Data Visualization and Analysis

Formulating analytical questions is a crucial step in guiding our exploration and analysis of the NYPD Shooting Incident Data.
I would like to categorize our questions in the following categories:

-   Temporal Trends
-   Geographical Trends
-   Demographical Trends
-   Incident Characteristics

We will go through them one by one and we would try to ask the most interesting questions in that area and see what the data can tell us about it.

## Temporal Trends

### Question

How have the number of shooting incidents varied by year ?

#### Visualization

```{r , echo=FALSE, fig.align='center'}
raw_data$OCCUR_YEAR <- as.numeric(format(as.Date(raw_data$OCCUR_DATE, format="%m/%d/%Y"), "%Y"))
# Summarize the data to get the count of incidents per year
yearly_incidents <- raw_data %>%
  group_by(OCCUR_YEAR) %>%
  summarise(Count = n())

# Create a bar plot with ggplot2
ggplot(yearly_incidents, aes(x = OCCUR_YEAR, y = Count)) +
  geom_bar(stat = "identity", fill = "gray") +
  theme_minimal() +
  labs(title = "Number of Shooting Incidents per Year", 
       x = "Year", 
       y = "Number of Incidents")

```

#### Analysis

There was a relatively high number of incidents in the early years shown on the graph, with a slight decrease leading up to around 2011.
Following this, there's a noticeable decline in the number of incidents for a few years.
However, after reaching a low, there is a marked increase in the number of incidents in recent years, with the number rising to levels similar to those seen in the earlier part of the timeline.This happened suddenly in 2020 (could it be related to any event or crisis happened at that time ? eg. Corona Virus?) We can observe an interesting decrease in the number of shooting between the year 2021 and 2022.

### Question

How have the number of shooting incidents varied by Month ?

#### Visualization

```{r , echo=FALSE, fig.align='center'}
raw_data$OCCUR_MONTH <- as.numeric(format(as.Date(raw_data$OCCUR_DATE, format="%m/%d/%Y"), "%m"))
raw_data$OCCUR_MONTH_NAME <- factor(raw_data$OCCUR_MONTH, levels = 1:12, labels = month.name)
# Summarize the data to get the count of incidents per month
monthly_incidents <- raw_data %>%
  group_by(OCCUR_MONTH_NAME) %>%
  summarise(Count = n())

# Create a bar plot with ggplot2
ggplot(monthly_incidents, aes(x = OCCUR_MONTH_NAME, y = Count)) + 
  geom_bar(stat = "identity", fill = "gray") +
  scale_x_discrete(name = "Month") +
  xlab("Month") +  # Label for the x-axis
  ylab("Number of Incidents") +  # Label for the y-axis
  ggtitle("Number of Shooting Incidents per Month") +  # Title of the plot
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate the x-axis labels to display vertically
```

#### Analysis

The number of incidents generally increases from January to a peak around July or August.
This could suggest a seasonal trend, with summer months experiencing higher incidents.
After reaching the peak, there is a decline in the number of incidents as the year progresses towards winter, with the lowest numbers typically in the early months of the year.
This pattern may indicate potential factors such as temperature changes, social behavior during warmer months, or other seasonal activities that could correlate with the frequency of shooting incidents.

## Geographical Trends

### Question

Which boroughs have the highest number of shooting incidents?

#### Visualization

```{r , echo=FALSE, fig.align='center'}
borough_counts <- raw_data %>%
  group_by(BORO) %>%
  summarise(Incidents = n()) %>%
  arrange(desc(Incidents))

ggplot(borough_counts, aes(x = reorder(BORO, Incidents), y = Incidents, fill = BORO)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Shooting Incidents by Borough",
       x = "Borough",
       y = "Number of Incidents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Analysis

The graph indicates a significant disparity in shooting incidents among New York City boroughs.
Brooklyn leads markedly, followed by the Bronx, with both significantly outpacing Queens, Manhattan, and particularly Staten Island, which has the fewest incidents.

### Question

Which precincts have the highest number of shooting incidents?

#### Visualization

```{r , echo=FALSE, fig.align='center'}
precinct_counts <- raw_data %>%
  group_by(PRECINCT) %>%
  summarise(Incidents = n()) %>%
  arrange(desc(Incidents))

# Since there are many precincts, we'll just take the top 10 for clarity
top_precincts <- head(precinct_counts, 10)
ggplot(top_precincts, aes(x = reorder(PRECINCT, Incidents), y = Incidents, fill = as.factor(PRECINCT))) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Precincts with the Highest Number of Shooting Incidents",
       x = "Precinct",
       y = "Number of Incidents") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Analysis

The bar chart shows the top 10 precincts with the highest number of shooting incidents.
The precinct labeled '75' has the highest count, indicating it's the most affected area.
The number of incidents gradually decreases across the precincts from left to right, with '73' and '67' also showing notably high counts.

### Question

Are there identifiable hotspots for shootings within the city?

#### Visualization

```{r , echo=FALSE, fig.align='center'}
raw_data$OCCUR_YEAR <- as.numeric(format(as.Date(raw_data$OCCUR_DATE, format="%m/%d/%Y"), "%Y"))
# Summarize the data to get the count of incidents per year

# Calculate the total incidents per precinct
total_incidents_per_precinct <- raw_data %>%
  group_by(PRECINCT) %>%
  summarise(TotalIncidents = n(), .groups = 'drop')

# Determine the top 10 precincts
top_precincts <- total_incidents_per_precinct %>%
  top_n(10, TotalIncidents) 

# Filter the data to include only the top 10 precincts
top_nypd_data <- raw_data %>%
  filter(PRECINCT %in% top_precincts$PRECINCT)

# Summarize the data to count incidents per precinct over the years for the top 10 precincts
incidents_per_precinct_year <- top_nypd_data %>%
  group_by(PRECINCT, OCCUR_YEAR) %>%
  summarise(Incidents = n(), .groups = 'drop') %>%
  arrange(PRECINCT, OCCUR_YEAR)

# Create a line plot with ggplot2
ggplot(incidents_per_precinct_year, aes(x = OCCUR_YEAR, y = Incidents, group = PRECINCT, color = as.factor(PRECINCT))) +
  geom_line() +
  labs(title = "Number of Incidents per Precinct Over the Years for Top 10 Precincts",
       x = "Year",
       y = "Number of Incidents") +
  theme_minimal() +
  theme(legend.position = "bottom")  # Place the legend at the bottom
```

#### Analysis

We can observe the variation of the number of incidents per Precinct per year.
The two constant hotspots which are almost always in the first two places are '75' and '73'.

## Demographical Trends

### Question

How do the ages of victims and perpetrators compare, and what age groups are most commonly involved in shooting incidents?

#### Visualization

```{r , echo=FALSE, fig.align='center'}

# Adjust the age group values
adjustted_nypd_data <- raw_data %>%
  mutate(PERP_AGE_GROUP = case_when(
    PERP_AGE_GROUP %in% c("940", "1020", "224") ~ "UNKNOWN",
    TRUE ~ as.character(PERP_AGE_GROUP)
  )) %>%
  mutate(VIC_AGE_GROUP = case_when(
    VIC_AGE_GROUP %in% c("940", "1020", "224") ~ "UNKNOWN",
    TRUE ~ as.character(VIC_AGE_GROUP)
  ))

# Summarize the data to count the number of incidents by age group for both victims and perpetrators
age_group_summary <- adjustted_nypd_data %>%
  group_by(PERP_AGE_GROUP, VIC_AGE_GROUP) %>%
  summarise(Incidents = n(), .groups = 'drop') %>%
  ungroup() %>%
  mutate(AgeGroup = ifelse(!is.na(PERP_AGE_GROUP), PERP_AGE_GROUP, VIC_AGE_GROUP),
         Role = ifelse(!is.na(PERP_AGE_GROUP), "Perpetrator", "Victim"))

# Create a side-by-side bar chart with ggplot2
ggplot(age_group_summary, aes(x = AgeGroup, y = Incidents, fill = Role)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Comparison of Age Groups between Victims and Perpetrators",
       x = "Age Group",
       y = "Number of Incidents") +
  scale_fill_manual(values = c("Perpetrator" = "blue", "Victim" = "red")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### Analysis

The bar chart compares the age groups of victims and perpetrators involved in shooting incidents.
The 25-44 age group is predominant for both roles, with perpetrators slightly outnumbering victims.
The least involved age group is 65+, while a considerable number of cases have unknown ages.
The age group \<18 has more victims than perpetrators represented.
It's also interesting that the number of perpetrators aged less then 18 is higher than the perpetrators in the age group 45-64.

### Question

What are the gender proportions among the victims and perpetrators?

#### Visualization

```{r , echo=FALSE, fig.align='center'}
# Ensure gender fields are treated as factors and filter out unknown or NA values
nypd_data <- raw_data %>%
  mutate(PERP_SEX = as.factor(PERP_SEX),
         VIC_SEX = as.factor(VIC_SEX)) %>%
  filter(PERP_SEX %in% c("M", "F"), VIC_SEX %in% c("M", "F"))

# Summarize the data to count the number of incidents by gender for victims and perpetrators
gender_summary <- nypd_data %>%
  count(PERP_SEX) %>%
  rename(Perpetrator_Count = n) %>%
  full_join(nypd_data %>% count(VIC_SEX) %>% rename(Victim_Count = n), by = c("PERP_SEX" = "VIC_SEX")) %>%
  pivot_longer(cols = c(Perpetrator_Count, Victim_Count), names_to = "Role", values_to = "Count") %>%
  mutate(Role = ifelse(Role == "Perpetrator_Count", "Perpetrator", "Victim"),
         PERP_SEX = ifelse(PERP_SEX == "M", "Male", "Female"))

# Create a bar plot with ggplot2
ggplot(gender_summary, aes(x = PERP_SEX, y = Count, fill = Role)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Gender Proportions among Victims and Perpetrators",
       x = "Gender",
       y = "Number of Incidents") +
  scale_fill_manual(values = c("Perpetrator" = "blue", "Victim" = "red")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### Analysis

The bar chart displays a stark contrast in gender proportions among victims and perpetrators in shooting incidents, with males significantly outnumbering females in both categories.
The disparity is particularly pronounced among perpetrators.

### Question

Are certain races or ethnicities overrepresented among the victims or perpetrators of shootings, and how do these patterns correlate with geographical locations within the city?

#### Visualization

```{r , echo=FALSE, fig.align='center'}
# Ensure race fields are treated as factors and filter out unknown or nonsensical values
nypd_data <- raw_data %>%
  filter(PERP_RACE %in% c("BLACK", "WHITE HISPANIC", "ASIAN / PACIFIC ISLANDER", "WHITE", "BLACK HISPANIC", "AMERICAN INDIAN/ALASKAN NATIVE"),
         VIC_RACE %in% c("BLACK", "WHITE HISPANIC", "ASIAN / PACIFIC ISLANDER", "WHITE", "BLACK HISPANIC", "AMERICAN INDIAN/ALASKAN NATIVE"))

# Summarize the data to count the number of incidents by race for victims and perpetrators
race_summary <- nypd_data %>%
  count(PERP_RACE) %>%
  rename(Perpetrator_Count = n) %>%
  full_join(nypd_data %>% count(VIC_RACE) %>% rename(Victim_Count = n), by = c("PERP_RACE" = "VIC_RACE")) %>%
  pivot_longer(cols = c(Perpetrator_Count, Victim_Count), names_to = "Role", values_to = "Count") %>%
  mutate(Role = ifelse(Role == "Perpetrator_Count", "Perpetrator", "Victim"))

# Create a bar plot with ggplot2
ggplot(race_summary, aes(x = PERP_RACE, y = Count, fill = Role)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Race and Ethnicity Proportions among Victims and Perpetrators",
       x = "Race/Ethnicity",
       y = "Number of Incidents") +
  scale_fill_manual(values = c("Perpetrator" = "blue", "Victim" = "red")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Analysis

The bar chart compares the racial and ethnic composition of victims and perpetrators in shooting incidents.
It shows a higher representation of Black individuals among both victims and perpetrators, followed by Hispanic and White individuals.
Asian/Pacific Islander and American Indian/Alaskan Native individuals have significantly lower counts in comparison.

# Modeling:

## Time Series forecasting using AutoRegressive Integrated Moving Average (ARIMA)

```{r , echo=FALSE, fig.align='center'}
# First, ensure the OCCUR_DATE is in Date format and extract the year and month
raw_data$YearMonth <- as.yearmon(as.Date(raw_data$OCCUR_DATE, format="%m/%d/%Y"))

# Aggregate the number of shootings per month
monthly_incidents <- raw_data %>%
  group_by(YearMonth) %>%
  summarise(Incidents = n(), .groups = 'drop')

# Convert to a time series
ts_data <- ts(monthly_incidents$Incidents, frequency = 12, start = c(year(min(raw_data$YearMonth)), month(min(raw_data$YearMonth))))

# Check for stationarity
adf.test(ts_data)

# If necessary, difference the data to make it stationary
ts_data_diff <- diff(ts_data, differences = 1)

# Model selection using ACF and PACF
Acf(ts_data_diff)
Pacf(ts_data_diff)

# Fit the ARIMA model
# The auto.arima function automatically selects the best order
arima_model <- auto.arima(ts_data)

# Check diagnostics
checkresiduals(arima_model)

# Forecast future values
forecasted_values <- forecast(arima_model, h = 12) # For example, forecasting the next 12 months

# Plot the forecast
plot(forecasted_values)
```
### Validation of the model:
the interpretation of the Ljung-Box test result provided:

Q\* = 45.104: This is the test statistic value.
It's compared against a chi-square distribution to determine the p-value.

df = 22: This indicates the degrees of freedom used in the test.
It represents the number of lags considered minus the number of parameters in the model.

p-value = 0.002575: The p-value is the probability of observing the test statistic as extreme as 45.104, under the null hypothesis that the residuals are independently distributed (no autocorrelation).

Since the p-value is less than the common significance levels (0.05, 0.01), we **reject** the null hypothesis, indicating that there is evidence of autocorrelation in the residuals at the lags up to 24.
This suggests that the ARIMA(1,0,0)(1,1,0)[12] with drift may not be a good fit as it's leaving some structure in the residuals that could be otherwise modeled.

## Time Series forecasting using a Negative Binomial model

```{r , echo=FALSE, fig.align='center'}
raw_data$Date <- as.Date(raw_data$OCCUR_DATE, format="%m/%d/%Y")
raw_data$Month <- floor_date(raw_data$Date, "month")

monthly_data <- raw_data %>%
  group_by(Month, PRECINCT) %>%
  summarise(Incident_Count = n(), .groups = 'drop')

# Fit a Negative Binomial model, predicting Incident_Count by Month and PRECINCT
nb_model <- glm.nb(Incident_Count ~ Month + as.factor(PRECINCT), data = monthly_data)

# Check the summary of the model
summary(nb_model)
```

### Validation of the model:

The provided summary output highlights various crucial aspects of the performance and adequacy of our Negative Binomial regression model:

1. **Significance Codes**: - 
These codes ( '\*\*\*', '\*\*', '\*', '\.', ' ') correspond to p-values of less than 0.001, 0.01, 0.05, 0.1, and 1, respectively. They serve as a convenient tool to evaluate the statistical significance of the predictors in our model. The greater the number of asterisks, the higher the level of statistical significance of the predictor.

2. **Dispersion Parameter**: - The dispersion parameter for the Negative Binomial family is specified as 6.1924, but in our model it is assumed to be 1. It is customary to use this approach when fitting Negative Binomial models in R. The value indicates the degree of overdispersion in our count data, where the variance is greater than the mean.

3. **Deviance**: - The Null Deviance (10827.3 on 8333 degrees of freedom) measures how well a model fits the data when no predictors, except for the intercept, are included. It quantifies the extent of the unexplained variation in the response variable by the model.
   The Residual Deviance (7155.4 on 8256 degrees of freedom) is lower than the null deviance, suggesting that the model with predictors is effectively explaining a substantial amount of the variability in the data. A lower residual deviance indicates a stronger fit between our model and the data.

The Akaike Information Criterion (AIC) is a statistical measure. The AIC value is 33910. The AIC (Akaike Information Criterion) is a useful metric for comparing models on the same dataset. A lower AIC value indicates a superior model. Nevertheless, AIC is subjective and particularly valuable when evaluating and contrasting various models.

5. **Theta (\\Theta)**: - Theta (6.192) represents the parameter of the Negative Binomial distribution that is associated with the variance. The standard error (0.252) offers an approximation of the extent of variation in this parameter estimate.

6. **Log-likelihood**: - The log-likelihood value (-33752.024) represents the probability of the observed data given the model that has been fitted. It is employed in the calculation of AIC and for the purpose of comparing models.

7. **Iterations of the Fisher Scoring algorithm**: - The convergence of the algorithm after only 1 iteration implies that the model was fitted in a direct and stable manner to the data.

### Explanation:

- our model demonstrates a notable enhancement compared to the ARIMA model, as evidenced by the decrease in deviance.
The user did not provide any text. The full summary output would indicate the statistical significance of individual predictors (not included here).
The user did not provide any text. The presence of overdispersion in our data, as indicated by the dispersion parameter, supports the utilization of a Negative Binomial model instead of a simpler Poisson model.
The user did not provide any text. The model's fit appears satisfactory, however, to conduct a thorough assessment, it is advisable to take into account the significance of individual predictors, diagnostic measures, and validation against other models or data partitions.

### Conclusion

In summary, the test results suggest that the ARIMA model as specified may not be fully adequate for the data, and further model exploration is needed.

# Identification of Potential Bias

We have explored different facets of shooting incident data, with a particular emphasis on temporal, geographical, demographic, and racial/ethnic patterns.
When evaluating potential biases in the dataset or analysis methods, several crucial factors emerge:

## Reporting Bias

Incidents may be subject to underreporting or overreporting in particular regions or among specific demographic cohorts as a result of disparities in police presence or community-police interactions.
The user did not provide any text.
The likelihood of victims or witnesses reporting incidents can vary based on their level of trust in law enforcement, potentially leading to a distortion in the data.

## Selection Bias

The dataset may lack incidents that were not reported to or recorded by the NYPD, resulting in an incomplete representation of all shooting incidents.
The user did not provide any text.
Unreported incidents may be influenced by various factors, including the gravity of the incident, the socio-economic status of the individuals involved, or the geographical location of the incident.

## Classification Bias

The racial and ethnic categorizations may be influenced by bias stemming from the reporting officer's subjective judgments or inconsistencies in how individuals involved in the incidents identify themselves.
- Age groups can be inaccurately documented or classified, particularly when identification is not accessible or when witnesses provide approximations.

## Geographical Bias

Some precincts may exhibit variations in data collection practices or discrepancies in the classification and reporting of incidents.
The user did not provide any text.
Certain regions may exhibit a higher degree of proactive crime reporting as a result of community initiatives, potentially leading to a greater number of reported incidents in comparison to other areas.

## Temporal Bias

Temporal Bias refers to the potential influence of changes in law enforcement practices, such as the adoption of new policies or technologies, on the reported number of shootings or the level of detail recorded about these incidents over time.
- Cultural and societal shifts, such as the increasing recognition and advocacy regarding gun violence, could impact the frequency of reporting and public opinions concerning such occurrences.

## Analysis Methodology

The selection of categories for analysis (such as racial groups or age brackets) may restrict the extent to which meaningful conclusions can be derived from the data.
The process of combining data into general categories can hide significant subtleties, such as variations within racial classifications or age cohorts.

## Interpretation Bias

The analyst's perspectives or assumptions may influence the conclusions drawn from the data, potentially resulting in an overemphasis or underrepresentation of certain trends.

Each of these biases has the potential to distort the genuine patterns and trends in the data, which could result in inaccurate conclusions or ineffective policy recommendations.
Recognizing these biases is crucial when interpreting the results, and, whenever feasible, employing methodologies that can alleviate their impact is imperative.
Some possible approaches to address this issue could involve employing statistical methodologies to account for established biases, cross-validating results with additional data sources, or explicitly acknowledging the limitations of the analysis.

# Conclusions

The key findings from the analysis of the NYPD Shooting Incident Data can be summarized as follows:

1.  **Temporal Trends**: Shooting incidents exhibit a seasonal pattern, with the highest occurrence observed in the summer months and the lowest in winter, indicating a potential association with seasonal influences.
2.  **Geographic Distribution**: Certain boroughs, such as Brooklyn and the Bronx, exhibit a greater concentration of incidents, suggesting distinct areas of high activity within the city.
3.  **Demographic Disparities**: Shooting incidents are predominantly observed among young adults, specifically those aged 25-44. Furthermore, there is a notable disparity in the likelihood of males being involved compared to females, with males being more frequently implicated as both perpetrators and victims.
4.  **Racial and Ethnic Representation**: The data indicates that Black individuals are disproportionately represented as both victims and perpetrators, followed by Hispanic and White individuals, while other racial groups have lower levels of representation.

### Consequences:

-   **Policy and Prevention**: Increased community involvement and proactive measures may be necessary during the summer season, when instances of gun violence are more prevalent.
-   **Optimal Resource Allocation**: Law enforcement resources can be more efficiently distributed by concentrating on identified hotspots within particular boroughs.
-   **Community Support**: Targeted interventions and support are necessary to address the root causes of young adult males' involvement in shootings, especially those from Black and Hispanic communities.
-   **Cultural Sensitivity**: Efforts to mitigate gun violence must demonstrate an awareness and understanding of diverse cultures, while also taking into account the intricate socio-economic elements that contribute to racial inequalities in shooting incidents.

### Recommendations for Further Investigation:

-   **Causal Factors**: Analyzing the root causes of seasonal fluctuations in shooting incidents in order to formulate targeted strategies for specific time periods.
-   **Efficacy of Interventions**: Evaluating the effectiveness of law enforcement strategies and community programs in mitigating the occurrence of shootings.
-   **Socio-economic Analysis**: Conducting an examination of the socio-economic conditions in areas with high shooting incidents in order to gain a comprehensive understanding of the larger context.
-   **Longitudinal Studies**: Performing longitudinal studies to observe the progression of trends over time and assess the enduring efficacy of various policies.
-   **Comparative Studies**: Analyzing New York City's data in relation to other cities in order to identify distinctive factors and effective strategies that can be replicated.

Future research should focus on mitigating the potential biases present in the dataset and utilizing rigorous statistical techniques to enhance the accuracy and representativeness of the findings.
By cross-referencing with qualitative data, such as community surveys or interviews, a more comprehensive understanding of the context and causes underlying the observed trends in the NYPD Shooting Incident Data can be obtained.
